Quantitative Estimates of Minimum Core Masses for Giant Planet Formation {\bf Paper I was quantitative.}

### 2014-03-03 tex/draft_paper2/draft_paper2.tex
minor Paper II edits, but more coming today

%Fast accretion implies large critical core masses. 
In order to calculate the minimum value of $M_{\rm crit}$, we first need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
%the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted. 
For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}). As a result, the core's envelope is in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.  As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences. In this case, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium. For a fixed planetesimal accretion rate and a set of disk conditions, $M_{\rm crit}$ is uniquely determined.

@@

\section{Model Review}
\label{sec2}

@@

We begin by reviewing the model developed in Paper I for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk. We describe the assumptions of the model and the properties of our assumed protoplanetary disk in \S\ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in \S\ref{struct}.

@@

We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective region and an outer radiative zone that matches smoothly onto the disk. The two regions are separated by the Schwarzschild criterion for convective instability (see \S\ref{struct}) . We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.  Moreover, we assume that the luminosity is constant throughout the outer radiative region (see Section \S\ref{critical} for additional discussion). We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction.The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance. We note that spherical symmetry confines us to the outer regions of the disk ($a\gtrsim5$ AU), where the disk scale height is larger than the radius at which the planet matches onto the disk (see Paper I for further details). The nebular gas is composed of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively. We assume that the envelope evolves through stages of quasi-static equilibrium.

@@

The temperature and pressure at the outer boundary of the atmosphere are given by the nebular temperature and pressure. As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. The surface density, mid-plane temperature and mid-plane pressure are given by

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200\, (a/\text{AU})^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120\, (a/\text{AU})^{-3/7} \,\,K \slabel{eq:diskb} \\
P\di&=&110\,  (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2} \slabel{eq:Pd} \, ,
\end{subeqnarray}

\noindent with $a$ the semimajor axis and for a mean molecular weight $\mu=2.35$.

@@

\noindent with $\sigma$ the Stefan-Boltzmann constant and $\kappa$ the dust opacity. In our models the atmosphere is optically thick throughout the outer boundary. Where energy is transported by convection, the temperature gradient is

\begin{equation}
\label{eq:delad}
\nabla = \delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\mathrm{ad}},
\end{equation}

\noindent with $\delad$ the adiabatic temperature gradient. The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla > \delad$. Since convective energy transport is highly efficient,  $\nabla \approx \delad$ in convecting regions. The temperature gradient is thus given by $\nabla=\mathrm{min}(\delad, \delrad)$.

@@

At low temperatures, hydrogen exists in molecular form and has a stable configuration. As the temperature becomes higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating. At $T \gtrsim 10,000$ K, hydrogen ionizes. In stellar and giant planet interiors there is little overlap between the two processes: hydrogen is almost entirely dissociated into atoms by the time ionization becomes important.

@@

An expression for the adiabatic gradient as a function of the dissociation fraction is presented in \App{deladdiss}. As expected, the adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen and $\delad=2/5$ when hydrogen is fully dissociated, but decreases significantly during partial dissociation and is smallest when half of the gas is dissociated. This is consistent with the behavior we see in Figure \ref{fig:deladmap}.

@@

Similarly to the case of dissociation, as $T$ increases, part of the rotational energy $U_{\rm r}$ of the hydrogen molecule is used to convert parahydrogen to orthohydrogen rather than to only increase the temperature, so $U_{\rm r} \propto T$ no longer holds. The adiabatic gradient $\delad$ thus varies with temperature. \App{deladspin} explores the temperature dependence in detail. These $\delad$ variations, which occur for $T \lesssim 300$ K, are displayed in Figure \ref{fig:deladmap}. At higher temperatures, the ortho-para 3:1 equilibrium ratio is reached; no further isomer conversion occurs, and thus $\delad$ remains relatively constant until dissociation temperatures are reached.

@@

So far in our model we have assumed that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity. However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling. Grain growth, in particular, changes the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}) and lowers the absolute value of the opacity. Enhanced metallicity due to planetesimal accretion, by construction not present during our atmosphere's growth, cannot make up for this reduction.

@@

\begin{equation}
\label{eq:graindistr}
n(a) \sim a^{-p},
\end{equation}
where $n(a)$ is the number density of particles of size $a$ and $p=3.5$ (corresponding to a ``normal'' collisional cascade) or $p=2.5$ (an approximation for coagulation). In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$. We take as a fiducial case a maximum particle size of 1 cm and a grain size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$. Other choices for the power-law coefficient $p$ are discussed later in this section.

The \citet{dalessio01} opacities are only relevant at temperatures that are sufficiently low for dust grains to remain solid ($T \lesssim 1000$ K). In the temperature regime where dust sublimates we use the \citet{bell94} analytic opacity laws, ensuring smooth transition from the grain growth opacities.

The sharp drop in opacity ($\kappa \sim T^{-24}$, see Figure \ref{fig:delvsr}) due to dust sublimation lowers the radiative temperature gradient significantly (see Equation \ref{eq:delrad}), and may therefore generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}). This could in principle pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid. In practice, however, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see Equation \ref{eq:structc}). This results in a negligible extra luminosity generated in the radiative windows. However, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section. We note that despite the existence of one or more radiative windows in the planet's atmosphere, the luminosity that emerges at the outer boundary is set by the outer radiative layer. In fact, a simple estimate of the luminosity emerging at the RCB can be found as follows.

@@

In this section we put together the results obtained in Sections \S\ref{EOSeffects}  and \S\ref{sec:opacity}, and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk, i.e. the critical core mass. As in Paper I, we quantify the runaway growth timescale $t_{\rm run}$ as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value (see Paper I for details). We first explore the dependence of the runaway growth time on the core mass for a fixed semimajor axis. We then determine the critical core mass to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas. Finally, we determine the critical core mass under more realistic opacity assumptions.

@@

Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs (e.g., \citealt{jay99}), for a gas described by a realistic EOS and an ISM dust opacity given by Equation (\ref{eq:opacitylaw}) with $F_{\kappa}=1$ and $\beta=2$. The results of Paper I for an ideal diatomic gas are plotted for comparison. The use of a realistic EOS results in a critical core mass more than twice as large as that produced by a polytrope. As such, non-ideal effects substantially increase the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.

@@

Figure \ref{fig:Mcritvsagg} shows the critical core mass as a function of semimajor axis, for a grain growth opacity with a size distribution given by the power-law (\ref{eq:graindistr}) with $p=3.5$ and maximum particle size of 1 cm. The critical core mass is lower than in the standard case, and less sensitive to location in the disk, i.e. the boundary conditions (temperature and pressure). We have shown in Paper I that the critical core mass is highly dependent on disk temperature rather than pressure. The temperature dependence, however, is mainly due to opacity. For the simplified analytic model developed in Paper I we found that the crossover time $t_{\rm co} \sim T_{\rm d}^{\beta+1/2}$, with $\beta$ the power-law exponent in Equation (\ref{eq:opacitylaw}) and $t_{\rm co}$ a proxy for $t_{\rm run}$, defined as the time when $M_{\rm atm}=M\co$, as $t_{\rm run}$ cannot be determined self-consistently for the analytic model. Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semimajor axis) dependence of the critical core mass, as seen in Figure \ref{fig:Mcritvsagg}. Moreover, grain growth reduces the absolute value of the opacity, which results in an overall lower runaway accretion time and critical core mass. % As noted in Section \S\ref{sec:opacity}, the critical core mass may be significantly lower if coagulation is taken into account, i.e. $p=2.5$.

@@

So far we have assumed that the size distribution of dust grains is a power-law (\ref{eq:graindistr}) with $p=3.5$ which corresponds to a normal collisional cascade. If, however, coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}. This results in a flatter and significantly lower opacity, which may substantially reduce the critical core mass. However, we have found that our model breaks down for low core masses ($M\co \lesssim 3 M_{\oplus}$) under our assumption of constant luminosity in the outer radiative layer. Figure \ref{fig:p25p35} shows the crossover time $t_{\rm co}$ as a function of semimajor axis for the lowest core mass for which our model is valid, $M\co=4 M_{\oplus}$. The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that the critical core mass may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.

@@

In this study we have considered protoplanets with fully formed cores for which planetesimal accretion is negligible and KH contraction dominates the luminosity evolution of the atmosphere. This is different from calculations that assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids. In this section we compare our results for the critical core mass to analogous results from steady-state fast planetesimal accretion calculations. We discuss the core accretion rates that are necessary for our regime to be valid in \S\ref{raf1}. In \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is negligible. Finally, we compare our results with those assuming fast planetesimal accretion in \S\ref{raf2}.

@@

\begin{equation}
\label{eq:McdotKH}
\dot{M\co}<\dot{M}_{\rm c, KH} \equiv \frac{L_{\rm KH} R\co}{G M\co}.
\end{equation}

To illustrate the magnitude of $\dot{M}_{\rm c, KH}$, we choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$. Since analytic studies of critical core masses at high planetesimal accretion rates assume an ideal gas EOS, for ease of comparison we choose an ideal gas polytrope with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also Paper I). For this choice of parameters, the runaway accretion time is $t _{\rm run}\sim$ 1.3 Myrs, which is within the typical lifetime of a protoplanetary disk. We also estimate two reference accretion rates. The first one is the core accretion rate $\dot{M}_{\rm c, acc}$ needed to grow the core to $M\co=10 M_{\oplus}$ on the same timescale as our model atmosphere, $\tau=1.3$ Myrs:

\begin{equation}
\label{eq:Mcdot}
\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \equiv \frac{M_{\rm{c}}}{\tau}.
\end{equation}
The second reference planetesimal accretion rate is $\dot{M}_{\rm c, Hill}$, a typically assumed planetesimal accretion rate for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core (for a review, see \citealt{goldreich04}). Following R06 (equation A1),

@@

Planetesimal accretion during the KH contraction phase of atmosphere growth at a rate $\dot{M}<\dot{M\co}_{\rm{KH}}$  cannot alter the core mass enough to affect the time evolution of the atmosphere. We can quantitatively estimate the increase in core mass as

@@

 \noindent from Equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model. For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.05 M_{\oplus} \ll 10 M_{\oplus}$. It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a much longer timescale then the crossover time for the unperturbed atmosphere. The time evolution of the atmosphere is thus insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.

@@

We compare our results for the critical core mass $M_{\rm crit}$ with those of studies that assume large planetesimal accretion rates. These studies find that the atmosphere is in steady state at all times and that static solutions only exist up to a maximum core mass, which they define as the critical core mass. If the $M_{\rm crit}$ found by static studies were lower than the $M_{\rm crit}$ in our regime of negligible solids accretion, the atmosphere would undergo runaway gas accretion before KH contraction became dominant, and our regime would not be relevant. We show, however, that our model yields lower core masses than those found when fast planetesimal accretion is considered.

@@

In order to estimate the critical core mass $M_{\rm crit, KH}$ corresponding to planetesimal accretion at the rate $\dot{M}_{\rm c, KH}$, we use the results of R06 for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest. R06 assumes an ideal gas polytropic EOS and a lower opacity than the standard ISM (see Equation \ref{eq:opacitylaw}). We thus calculate the critical core mass for an ideal gas polytrope with the normalization factor $F_{\kappa}$ reduced by a hundred, which is comparable to the opacity law used by R06 \footnote{The power-law opacity of R06 is scaled to the (semimajor axis dependent) disk temperature, while our opacity is scaled to an absolute reference temperature. We thus cannot directly use the R06 opacities for our comparison.}. %(see Paper I).

@@

\begin{equation}
\label{eq:u}
U=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V}
\end{equation}
\begin{equation}
\label{eq:s}
S=\mathcal{R} \log{Z} + \frac{U}{T}
\end{equation}
\begin{equation}
\label{eq:cv}
C_v=\Big(\frac{\partial U}{\partial T}\Big)_{V}.
\end{equation}
Since $Z=Z_t Z_r Z_v$, we may write $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and vibration partition functions, respectively.

@@

\begin{equation}
\label{eq:ut}
U_t=\frac{3}{2} \mathcal{R} T
\end{equation}

@@

\item{\textbf{Density.}} In the low temperature, low pressure regime, hydrogen is molecular and behaves like an ideal gas. As such, the density in this region follows the ideal gas law $P=\rho \mathcal{R} T$.
\item{\textbf{Internal energy per mass.}} $U=U_t+U_r+U_v$, where $U_t$ is given by Equation(\ref{eq:ut}), and $U_r$ and $U_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
\item{\textbf{Entropy per unit mass}}. Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by Equation (\ref{eq:st}), and $S_r$ and $S_v$ can be determined from Equation (\ref{eq:s}) and the calculated expressions for $U_r$ and $U_v$, respectively.
\item{\textbf{Entropy logarithmic derivatives}}. The logarithmic derivatives $S_T$ and $S_P$ are given by:

>>>>

%Fast accretion implies large critical core masses. 
In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
%the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted. 
For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}). As a result, the core's envelope is in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.  As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences. In this case, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium. For a fixed planetesimal accretion rate and a set of disk conditions, $M_{\rm crit}$ is uniquely determined.

@@

\section{Atmospheric Model Review}
\label{sec2}

@@

In this section we review the model developed in Paper I for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk. We describe the assumptions of the model and the properties of our assumed protoplanetary disk in \S\ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in \S\ref{struct}.

@@

We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective region and an outer radiative zone that matches smoothly onto the disk. The two regions are separated by the Schwarzschild criterion for convective instability (see \S\ref{struct}). The surface between the two regions is the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.  Moreover, we assume that the luminosity is constant throughout the radiative region (see Section \S\ref{critical} for additional discussion). We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction.The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance. We note that spherical symmetry confines us to the outer regions of the disk ($a\gtrsim5$ AU), where the disk scale height is larger than the radius at which the planet matches onto the disk (see Paper I for further details). The nebular gas is composed of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively. We assume that the envelope evolves through stages of quasi-static equilibrium.

@@

The temperature and pressure at the outer boundary of the atmosphere are given by the nebular temperature and pressure. We use the minimum mass, passively irradiated disk model of  \citet{chiang10}. The surface density, mid-plane temperature and mid-plane pressure are

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200\, (a/\text{AU})^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120\, (a/\text{AU})^{-3/7} \,\,K \slabel{eq:diskb} \\
P\di&=&110\,  (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2} \slabel{eq:Pd} \, ,
\end{subeqnarray}
where $a$ is the semimajor axis and for a mean molecular weight $\mu=2.35$.

@@

\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the dust opacity. In our models the atmosphere is optically thick throughout the outer boundary. Where energy is transported by convection, the temperature gradient is

\begin{equation}
\label{eq:delad}
\nabla = \delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\mathrm{ad}},
\end{equation}
with $\delad$ the adiabatic temperature gradient. The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla > \delad$. Since convective energy transport is highly efficient,  $\nabla \approx \delad$ in convecting regions. The temperature gradient is thus given by $\nabla=\mathrm{min}(\delad, \delrad)$.

@@

At low temperatures, hydrogen is molecular and has a stable configuration. At temperatures higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating. At $T \gtrsim 10,000$ K, hydrogen ionizes. In stellar and giant planet interiors there is little overlap between the two processes: hydrogen is almost entirely dissociated into atoms by the time ionization becomes important.

@@

An expression for the adiabatic gradient as a function of the dissociation fraction is presented in \App{deladdiss}. As expected, the adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen and $\delad=2/5$ when hydrogen is fully dissociated, but decreases significantly during partial dissociation and is smallest when half of the gas is dissociated. This is consistent with the behavior shown in Figure \ref{fig:deladmap}.

@@

Similarly to the case of dissociation, as $T$ increases, part of the rotational energy of the hydrogen molecule is used to convert parahydrogen to orthohydrogen rather than to only increase the temperature. Thus $u_{\rm r} \propto T$, where $u_{\rm r}$ is the rotational energy per unit mass, no longer holds, and the adiabatic gradient $\delad$ is temperature dependent. These $\delad$ variations, which occur for $T \lesssim 300$ K, are explored in \App{deladspin} and displayed in Figure \ref{fig:deladmap}. At higher temperatures, the ortho-para 3:1 equilibrium ratio is reached; no further isomer conversion occurs, and thus $\delad$ remains relatively constant until dissociation temperatures are reached.

@@

Our calculations so far have assumed that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity. However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling. Grain growth, in particular, changes the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}) and lowers the absolute value of the opacity. Enhanced metallicity due to planetesimal accretion, by construction not present during our atmosphere's growth, cannot make up for this reduction.

@@

\begin{equation}
\label{eq:graindistr}
n(a) \sim a^{-p},
\end{equation}
where $n(a)$ is the number density of particles of size $a$ and $p=3.5$ (corresponding to a ``normal'' collisional cascade) or $p=2.5$ (an approximation for coagulation). In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$. We take as a fiducial case a maximum particle size $a_{\rm max}=1$ cm and a grain size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$. Other choices for the power-law coefficient $p$ are discussed later in this section.

The \citet{dalessio01} opacities are only relevant at temperatures that are sufficiently low for dust grains to remain solid ($T \lesssim 1000$ K). At higher temperatures, we use the \citet{bell94} analytic opacity laws, ensuring smooth transition from the grain growth opacities.

The sharp drop in opacity ($\kappa \sim T^{-24}$, see Figure \ref{fig:delvsr}) due to dust sublimation lowers the radiative temperature gradient significantly (see Equation \ref{eq:delrad}), and may thus generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}). This could in principle pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid. In practice, however, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see Equation \ref{eq:structc}). This results in a negligible extra luminosity generated in the radiative windows. However, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section. Despite the existence of one or more radiative windows in the planet's atmosphere, the luminosity that emerges at the outer boundary is set by the outer radiative layer. A simple estimate of the luminosity emerging at the RCB can be found as follows.

@@

In this section we put together the results obtained in Sections \S\ref{EOSeffects}  and \S\ref{sec:opacity}, and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk, i.e. the critical core mass. As in Paper I, we quantify the runaway growth timescale $t_{\rm run}$ as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value (see Paper I for details). We first explore the dependence of the runaway growth time on the core mass for a fixed semimajor axis. We then determine the critical core mass $M_{\rm crit}$ to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas. Finally, we determine $M_{\rm crit}$ under more realistic opacity assumptions.

@@

Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs (e.g., \citealt{jay99}), for a gas described by a realistic EOS and an ISM dust opacity given by Equation (\ref{eq:opacitylaw}) with $F_{\kappa}=1$ and $\beta=2$. The results of Paper I for an ideal diatomic gas are plotted for comparison. The inclusion of realistic EOS effects increases $M_{\rm crit}$ by more than a factor of two when compared to an ideal gas polytrope. As such, non-ideal effects substantially increase the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.

@@

Figure \ref{fig:Mcritvsagg} shows $M_{\rm crit}$ as a function of semimajor axis, for a grain growth opacity with a size distribution given by the power-law (\ref{eq:graindistr}) with $p=3.5$ and maximum particle size $a_{\rm max}=1$ cm. The critical core mass is lower than in the standard interstellar opacity case, and less sensitive to location in the disk, i.e. the boundary conditions (temperature and pressure). We have shown in Paper I that the critical core mass is highly dependent on disk temperature rather than pressure. The temperature dependence, however, is mainly due to opacity. For the simplified analytic model developed in Paper I we found that the crossover time $t_{\rm co} \sim T_{\rm d}^{\beta+1/2}$, with $\beta$ the power-law exponent in Equation (\ref{eq:opacitylaw}) and $t_{\rm co}$ a proxy for $t_{\rm run}$, defined as the time when $M_{\rm atm}=M\co$, as $t_{\rm run}$ cannot be determined self-consistently for the analytic model. Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semimajor axis) dependence of the critical core mass, as seen in Figure \ref{fig:Mcritvsagg}. Moreover, grain growth reduces the absolute value of the opacity, which results in an overall lower runaway accretion time and critical core mass. % As noted in Section \S\ref{sec:opacity}, the critical core mass may be significantly lower if coagulation is taken into account, i.e. $p=2.5$.

@@

In the grain size distribution (\ref{eq:graindistr}), the coefficient $p=3.5$ corresponds to a normal collisional cascade. If coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}. This results in a flatter and significantly lower opacity, which may substantially reduce $M_{\rm crit}$. However, we have found that our model breaks down for low core masses ($M\co \lesssim 3 M_{\oplus}$) under our assumption of constant luminosity in the outer radiative layer. Figure \ref{fig:p25p35} shows the crossover time $t_{\rm co}$ as a function of semimajor axis for the lowest core mass for which our model is valid, $M\co=4 M_{\oplus}$. The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that $M_{\rm crit}$ may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.

@@

This study considers protoplanets with fully formed cores for which planetesimal accretion is negligible and KH contraction dominates the luminosity evolution of the atmosphere. This is different from calculations that assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids. In this section we compare our results for the critical core mass to analogous results from steady-state fast planetesimal accretion calculations. We discuss the core accretion rates that are necessary for our regime to be valid in \S\ref{raf1}. In \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is indeed negligible. Finally, we compare our results with those assuming fast planetesimal accretion in \S\ref{raf2}.

@@

\begin{equation}
\label{eq:McdotKH}
\dot{M\co}<\dot{M}_{\rm c, KH} \equiv \frac{L_{\rm KH} R\co}{G M\co}.
\end{equation} 
To illustrate the magnitude of $\dot{M}_{\rm c, KH}$, we choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$. Since analytic studies of critical core masses at high planetesimal accretion rates assume an ideal gas EOS, for ease of comparison we choose an ideal gas polytropic EOS with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also Paper I). For this choice of parameters, the runaway accretion time is $t _{\rm run}\sim$ 1.3 Myrs, which is within the typical lifetime of a protoplanetary disk. We also estimate two reference accretion rates. The first one is the core accretion rate $\dot{M}_{\rm c, acc}$ needed to grow the core to $M\co=10 M_{\oplus}$ on the same timescale as our model atmosphere, $\tau=1.3$ Myrs:

\begin{equation}
\label{eq:Mcdot}
\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \sim \frac{M_{\rm{c}}}{\tau}.
\end{equation}
The second reference planetesimal accretion rate is $\dot{M}_{\rm c, Hill}$, a typically assumed planetesimal accretion rate for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core (for a review, see \citealt{goldreich04}). Following R06 (equation A1),

@@

Planetesimal accretion during the KH contraction phase of atmosphere growth at a rate $\dot{M}_{\rm c}<\dot{M}_{\rm{c,KH}}$  cannot alter the core mass enough to affect the time evolution of the atmosphere. We can quantitatively estimate the increase in core mass as

@@

 \noindent from Equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model. For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.05 M_{\oplus} \ll 10 M_{\oplus}$. Core growth is negligible in our regime, and the time evolution of the atmosphere is thus insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.

@@

We compare our results for $M_{\rm crit}$ with those of studies that assume large planetesimal accretion rates. These studies find that the atmosphere is in steady state at all times and that static solutions only exist up to a maximum (critical) core mass. If the $M_{\rm crit}$ found by static studies were lower than the $M_{\rm crit}$ in our regime of negligible solids accretion, the atmosphere would undergo runaway gas accretion before KH contraction became dominant, and our regime would not be relevant. We show, however, that our model yields lower core masses than those found when fast planetesimal accretion is considered.

@@

In order to estimate the critical core mass $M_{\rm crit, KH}$ corresponding to planetesimal accretion at the rate $\dot{M}_{\rm c, KH}$, we use the results of R06 for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest. R06 assumes an ideal gas polytropic EOS and a lower opacity than the standard ISM (see Equation \ref{eq:opacitylaw}). We thus calculate $M_{\rm crit}$ for an ideal gas polytrope with the normalization factor $F_{\kappa}$ reduced by a hundred, which is comparable to the opacity law used by R06 \footnote{The power-law opacity of R06 is scaled to the (semimajor axis dependent) disk temperature, while our opacity is scaled to an absolute reference temperature. We thus cannot directly use the R06 opacities for our comparison.}. %(see Paper I).

@@

\begin{equation}
\label{eq:u}
u=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V}
\end{equation}
\begin{equation}
\label{eq:s}
S=\mathcal{R} \log{Z} + \frac{U}{T}
\end{equation}
\begin{equation}
\label{eq:cv}
C_v=\Big(\frac{\partial U}{\partial T}\Big)_{V}.
\end{equation}
Since $Z=Z_t Z_r Z_v$, we may write $u=u_t+u_r+u_v$ and $S=S_t+S_r+S_v$, where $u_t$, $u_r$, $u_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and vibration partition functions, respectively.

@@

\begin{equation}
\label{eq:ut}
u_t=\frac{3}{2} \mathcal{R} T
\end{equation}

@@

\item{\textbf{Density.}} In the low temperature, low pressure regime, hydrogen is molecular and behaves like an ideal gas. As such, the density in this region follows the ideal gas law $P=\rho \mathcal{R} T$.
\item{\textbf{Internal energy per mass.}} $u=u_t+u_r+u_v$, where $u_t$ is given by Equation(\ref{eq:ut}), and $u_r$ and $u_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
\item{\textbf{Entropy per unit mass}}. Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by Equation (\ref{eq:st}), and $S_r$ and $S_v$ can be determined from Equation (\ref{eq:s}) and the calculated expressions for $u_r$ and $u_v$, respectively.
\item{\textbf{Entropy logarithmic derivatives}}. The logarithmic derivatives $S_T$ and $S_P$ are given by:
